Implementation of Parallel Algorithms for ARIMA in Distributed Database Systems

### 2013-09-23 draft.tex
Update

\documentclass[english,12pt]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{float} \usepackage{graphicx}
\usepackage{esint} \usepackage{natbib} \usepackage{times,bm} \usepackage{babel}
\usepackage{amsmath}
\usepackage{algorithm2e}

\makeatletter
\linespread{1.25}
\pagestyle{headings}

@@

@@

% Introduce to MADlib, and implement ARIMA in MADLib
MADlib \cite{madlib} is an open source library for scalable in-database
analytics.  It is created by the Predictive Analytics Team at Pivotal Inc.
MADlib provides data-parallel implementations of mathematical, statistical and
machine-learning methods for structured and unstructured data.

@@

Cohen et al. \cite{mad-skills} explained that "MAD" stands for "magnetic",
"agile", "deep", and "lib" stands for advanced (mathematical, statistical,
machine learning), parallel and scalable in-database functions.

The MADlib project was initiated in late 2010. Currently MADlib's latest
version is 1.2. By itself MADlib provides a pure SQL interface. An R front-end
named PivotalR \cite{pivotalr} is also created by the Predictive Analytics Team
at Pivotal Inc. A Python wrapper \cite{python-madlib} for MADlib is also
available.

@@

ARIMA models involve the following variables:
\begin{enumerate}
   \item The lag difference $Y_{t}$, where  $Y_{t} = (1-B)^{d}(X_{t} - \mu)$.
    \item The values of the time series $X_t$.
    \item $p$, $q$, and $d$ are the parameters of the ARIMA model.
      $d$ is the differencing order, $p$ is the order of the AR
      operator, and $q$ is the order of the MA operator.
    \item The AR operator $\phi(B)$.
    \item The MA operator $\theta(B)$.
    \item The mean value $\mu$, which is always set to be zero for
      $d>0$ or need to be estimated.
    \item The error terms $Z_t$.
\end{enumerate}

The  auto regression operator models the prediction for the next
observation  as some linear combination of the previous observations.
More formally, an AR operator of order $p$ is defined as
\begin{align}
\phi(B) Y_t= \phi_1 Y_{t-1}   + \dots +  \phi_{p} Y_{t-p}
\end{align}

The moving average operator is similar, and it models the prediction
for the next observation as a linear combination of the errors in the
previous prediction errors.  More formally, the MA operator of order
$q$ is defined as
\begin{align}
\theta(B) Z_t =   \theta_{1} Z_{t-1} + \dots + \theta_{q} Z_{t-q}.
\end{align}

We assume that
\begin{equation}
\Pr(Z_t) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-Z^2_t/2 \sigma^2}, \quad t > 0
\end{equation}
and that  $Z_{-q+1} = Z_{-q+2} = \dots = Z_0 = Z_1 = \dots = Z_p =
0$. The initial values of $Y_t=X_t-\mu$ for $t=-p+1, -p+2, \dots,
0$ can be solved from the following linear equations
\begin{eqnarray}
\phi_1 Y_0 + \phi_2 Y_{-1} + \cdots + \phi_p Y_{-p+1} &=& Y_1 \nonumber\\
\phi_2 Y_0 + \cdots + \phi_p Y_{-p+2} &=& Y_2 - \phi_1 Y_1  \nonumber\\
&\vdots& \nonumber\\
\phi_{p-1} Y_0 + \phi_p Y_{-1} &=& Y_{p-1} - \phi_1 Y_{p-2} - \cdots -
\phi_{p-2} Y_1 \nonumber \\
\phi_p Y_0  &=& Y_p - \phi_1 Y_{p-1} - \cdots - \phi_{p-1} Y_{1} \label{eq:init_Y}
\end{eqnarray}

The likelihood function $L$ for $N$ values of $Z_t$  is then
\begin{equation}
L(\phi, \theta) = \prod_{t = 1}^N  \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-Z^2_t/2 \sigma^2}
\end{equation}
so the log likelihood function $l$ is
\begin{align}
l(\phi, \theta) &= \sum_{t = 1}^N \ln \left(\frac{1}{\sqrt{2 \pi \sigma^2}} e^{-Z^2_t/2 \sigma^2}
 \right) \nonumber\\
 &=  \sum_{t = 1}^N  - \ln \left( \sqrt{2 \pi \sigma^2}\right)  -\frac{Z^2_t}{2 \sigma^2}\nonumber\\
&=  -\frac{N}{2} \ln \left( 2 \pi \sigma^2\right)  - \frac{1}{2
  \sigma^2} \sum_{t = 1}^N   Z^2_t\ \. \label{eq:loglikelihood}
\end{align}
Thus, finding the maximum likelihood is equivalent to solving the
optimization problem (known as the conditional least squares
formation)
\begin{align}
\min_{\theta, \phi} \sum_{t = 1}^N  Z^2_t.
\end{align}
The error term $Z_t$ can be computed iteratively as follows:
\begin{align}\label{equ:error-terms}
Z_t = X_t - F_t(\phi, \theta, \mu)
\end{align}
where
\begin{align}
F_t(\phi, \theta, \mu) = \mu + \sum_{i=1}^p \phi_i (X_{t-i}-\mu) + \sum_{i=1}^q \theta_i Z_{t-i}
\end{align}

@@

%Author's (Mark W) Note: This is taken from http://people.duke.edu/~hpgavin/ce281/lm.pdf
To understand the Levenberg-Marquardt algorithm, it helps to know the
gradient descent method and the Gauss-Newton method.  On many
``reasonable'' functions, the gradient descent method takes large
steps when the current iterate is distant from the true solution, but
is slow to converge an the current iterate nears the true solution.
The Gauss-Newton method is much faster for converging when the current
iterate is in the neighborhood of the true solution.  The
Levenberg-Marquardt algorithm tries to get the best of best worlds,
and combine the gradient descent step with Gauss-Newton step in a
weighted average.  For iterates far from the true solution, the step
favors the gradient descent step, but as the iterate approaches the
true solution, the Gauss-Newton step dominates.

%Author's (Mark W) Note: Sudo code taken from http://users.ics.forth.gr/~lourakis/levmar/levmar.pdf
Like other numeric minimization algorithms, LMA is an iterative
procedure.  To start a minimization, the user has to provide an
initial guess for the parameter vector, $p$, as well as some tuning
parameters $\tau, \epsilon_1, \epsilon_2, \epsilon_3,$ and $k_{max}$.
Let $Z(p)$ be the vector of calculated errors ($Z_t$'s) for the
parameter vector $p$, and let $J = (J_{1}, J_{2}, \dots, J_N)^T$
be a Jacobian matrix.

@@

\begin{algorithm}[H]
    \SetAlgoLined
\alginput{An initial guess for parameters $\vec{\phi}_0, \vec{\theta}_0, \mu_0$}
\algoutput{The parameters that maximize the likelihood $\vec{\phi}^*,
  \vec{\theta}^*, \mu^*$}
\begin{algorithmic}[1]
	\State $k \leftarrow 0$  \Comment{Iteration counter}
	\State $v \leftarrow 2$ \Comment{The change in the weighting factor.}
	\State $(\vec{\phi},\vec{\theta},\mu) \leftarrow (\vec{\phi}_0,\vec{\theta}_0,\mu_0)$ \Comment{Initialize parameter vector}
    \State Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
    equation~\ref{equ:error terms}.  \Comment{Vector of errors}
	\State $A \leftarrow J^T J$   \Comment{The  Gauss-Newton Hessian approximation}
	\State $u \leftarrow \tau * \max_i(A_{ii})$ \Comment{Weight of the gradient-descent step}
	\State $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$ \Comment{The gradient descent step.}
	\State $ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$ \Comment{Termination Variable}
	\While{(not stop) and ($k < k_{max}$)}
		\State $k \leftarrow k + 1$
		\Repeat
			\State $\delta \leftarrow (A + u \times \text{diag}(A))^{-1} g$ \Comment{Calculate step direction}
			\If{$\| \delta \| \le \epsilon_2 \|
              (\vec{\phi},\vec{\theta},\mu) \|$} \Comment{Change in the parameters is too small to continue.}
				\State  $\text{stop} \leftarrow \text{true}$
			\Else
				\State $(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new}) \leftarrow (\vec{\phi},\vec{\theta},\mu) + \delta$ \Comment{Take a trial step in the new direction}
				\State $\rho \leftarrow (\| Z(\vec{\phi},\vec{\theta},\mu)\|^2 - \| Z(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})\|^2 )/(\delta^T(u \delta + g))$ \Comment{Calculate improvement of trial step}
				\If{$\rho > 0$} \Comment{Trial step was good, proceed to next iteration}
					\State $(\vec{\phi},\vec{\theta},\mu) \leftarrow (\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})$ \Comment{Update variables}
                    \State Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
                    equation~\ref{equ:error terms}.
					\State $A \leftarrow J^T J$
					\State $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$
					\State $ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$ or $(\| Z(\vec{\phi},\vec{\theta},\mu)^2 \| \le \epsilon_3)$  \Comment{Terminate if we are close to the solution.}
					\State $v \leftarrow 2$
					\State $u \rightarrow u * \max(1/3, 1 - (2\rho - 1)^3 )$
				\Else  \Comment{Trial step was bad, change weighting on the gradient decent step}
					\State $v \leftarrow 2 v$
					\State $u \leftarrow u v$
				\EndIf
			\EndIf
		\Until{(stop) or ($\rho > 0$) }
	\EndWhile
	\State $(\vec{\phi}^*,\vec{\theta}^*,\mu^*) \leftarrow (\vec{\phi},\vec{\theta},\mu)$
\end{algorithmic}
\label{alg:Levenberg-Marquardt}
\end{algorithm}

Suggested values for the tuning parameters are $\epsilon_1 = \epsilon_2 =
\epsilon_3 = 10^{-15}, \tau = 10^{-3}$ and $k_{max} = 100$.

@@

% What are the difficulties that we are facing

>>>>

\documentclass[english,12pt]{article}
\usepackage{babel}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{algorithm2e}

%\makeatletter
\linespread{1.25}

% \pagestyle{headings}

@@

%\maketitle

@@

% Introduce to MADlib, and implement ARIMA in MADLib
MADlib~\cite{madlib} is an open source library for scalable in-database
analytics.  It is created by the Predictive Analytics Team at Pivotal Inc.
MADlib provides data-parallel implementations of mathematical, statistical and
machine-learning methods for structured and unstructured data.

@@

Cohen et al.~\cite{mad-skills} explained that ``MAD'' stands for ``magnetic'',
``agile'', ``deep'', and ``lib'' stands for advanced (mathematical, statistical,
machine learning), parallel and scalable in-database functions.

The MADlib project was initiated in late 2010. Currently MADlib's latest
version is 1.2. By itself MADlib provides a pure SQL interface. An R front-end
named PivotalR~\cite{pivotalr} is also created by the Predictive Analytics Team
at Pivotal Inc. A Python wrapper~\cite{python-madlib} for MADlib is also
available.

@@

ARIMA models involve the following variables:
\begin{enumerate}
   \item The lag difference $Y_{t}$, where  $Y_{t} = {(1-B)}^{d}(X_{t} - \mu)$.
    \item The values of the time series $X_t$.
    \item $p$, $q$, and $d$ are the parameters of the ARIMA model.
      $d$ is the differencing order, $p$ is the order of the AR
      operator, and $q$ is the order of the MA operator.
    \item The AR operator $\phi(B)$.
    \item The MA operator $\theta(B)$.
    \item The mean value $\mu$, which is always set to be zero for
      $d>0$ or need to be estimated.
    \item The error terms $Z_t$.
\end{enumerate}

The  auto regression operator models the prediction for the next
observation  as some linear combination of the previous observations.
More formally, an AR operator of order $p$ is defined as
\begin{equation}
\phi(B) Y_t= \phi_1 Y_{t-1}   + \cdots +  \phi_{p} Y_{t-p}
\end{equation}

The moving average operator is similar, and it models the prediction
for the next observation as a linear combination of the errors in the
previous prediction errors.  More formally, the MA operator of order
$q$ is defined as
\begin{equation}
\theta(B) Z_t =   \theta_{1} Z_{t-1} + \cdots + \theta_{q} Z_{t-q}.
\end{equation}

We assume that
\begin{equation}
\Pr(Z_t) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-Z^2_t/2 \sigma^2}, \quad t > 0
\end{equation}
and that  $Z_{-q+1} = Z_{-q+2} = \cdots = Z_0 = Z_1 = \cdots = Z_p =
0$. The initial values of $Y_t=X_t-\mu$ for $t=-p+1, -p+2, \dots,
0$ can be solved from the following linear equations
\begin{eqnarray}
\phi_1 Y_0 + \phi_2 Y_{-1} + \cdots + \phi_p Y_{-p+1} &=& Y_1 \nonumber\\
\phi_2 Y_0 + \cdots + \phi_p Y_{-p+2} &=& Y_2 - \phi_1 Y_1  \nonumber\\
&\vdots& \nonumber\\
\phi_{p-1} Y_0 + \phi_p Y_{-1} &=& Y_{p-1} - \phi_1 Y_{p-2} - \cdots -
\phi_{p-2} Y_1 \nonumber \\
\phi_p Y_0  &=& Y_p - \phi_1 Y_{p-1} - \cdots - \phi_{p-1} Y_{1} \label{eq:init_Y}
\end{eqnarray}

The likelihood function $L$ for $N$ values of $Z_t$  is then
\begin{equation}
L(\phi, \theta) = \prod_{t = 1}^N  \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-Z^2_t/2 \sigma^2}
\end{equation}
so the log likelihood function $l$ is
\begin{eqnarray}
l(\phi, \theta) &=& \sum_{t = 1}^N \ln \left(\frac{1}{\sqrt{2 \pi \sigma^2}}
 e^{-Z^2_t/2 \sigma^2}
 \right) \nonumber\\
 &=&  \sum_{t = 1}^N  - \ln \left( \sqrt{2 \pi \sigma^2}\right)
-\frac{Z^2_t}{2 \sigma^2}\nonumber\\
&=&  -\frac{N}{2} \ln \left( 2 \pi \sigma^2\right)  - \frac{1}{2
  \sigma^2} \sum_{t = 1}^N   Z^2_t\ . \label{eq:loglikelihood}
\end{eqnarray}
Thus, finding the maximum likelihood is equivalent to solving the
optimization problem (known as the conditional least squares
formation)
\begin{equation}
\min_{\theta, \phi} \sum_{t = 1}^N  Z^2_t.
\end{equation}
The error term $Z_t$ can be computed iteratively as follows:
\begin{equation}
    Z_t = X_t - F_t(\phi, \theta, \mu) \label{eq:error-terms}
\end{equation}
where
\begin{equation}
F_t(\phi, \theta, \mu) = \mu + \sum_{i=1}^p \phi_i (X_{t-i}-\mu) + \sum_{i=1}^q
\theta_i Z_{t-i}
\end{equation}

@@

To understand the Levenberg-Marquardt algorithm, it helps to know the
gradient descent method and the Gauss-Newton method.  On many
``reasonable'' functions, the gradient descent method takes large
steps when the current iterate is distant from the true solution, but
is slow to converge an the current iterate nears the true solution.
The Gauss-Newton method is much faster for converging when the current
iterate is in the neighborhood of the true solution.  The
Levenberg-Marquardt algorithm tries to get the best of best worlds,
and combine the gradient descent step with Gauss-Newton step in a
weighted average.  For iterates far from the true solution, the step
favors the gradient descent step, but as the iterate approaches the
true solution, the Gauss-Newton step dominates.

Like other numeric minimization algorithms, LMA is an iterative
procedure.  To start a minimization, the user has to provide an
initial guess for the parameter vector, $p$, as well as some tuning
parameters $\tau$, $\epsilon_1$, $\epsilon_2$, $\epsilon_3$, and $k_{\max}$.
Let $Z(p)$ be the vector of calculated errors ($Z_t$'s) for the
parameter vector $p$, and let $J = {(J_{1}, J_{2}, \ldots, J_N)}^T$
be a Jacobian matrix.

@@

\begin{algorithm}[ht]
\KwIn{An initial guess for parameters $\vec{\phi}_0, \vec{\theta}_0, \mu_0$}
\KwOut{The parameters that maximize the likelihood $\vec{\phi}^*,
  \vec{\theta}^*, \mu^*$}
    $k \leftarrow 0$; $v \leftarrow 2$;
    $(\vec{\phi},\vec{\theta},\mu) \leftarrow
    (\vec{\phi}_0,\vec{\theta}_0,\mu_0)$\;
    Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
    Eq. (9).  \textit{\footnotesize \# Vector of errors}\;
    $A \leftarrow J^T J$   \textit{\footnotesize \# The Gauss-Newton Hessian
    approximation}\;
    $u \leftarrow \tau * \max_i(A_{ii})$ \textit{\footnotesize \# Weight of the
    gradient-descent step}\;
    $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$ \textit{\footnotesize \#
    The gradient descent step}\;
	$ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$
    \textit{\footnotesize \# Termination Variable}\;
	\While{not stop and $k < k_{\max}$}{
		$k \leftarrow k + 1$\;
		\Repeat{stop or $\rho > 0$}{
			$\delta \leftarrow {(A + u \times \text{diag}(A))}^{-1} g$
            \textit{\footnotesize \# Calculate step direction}\;
            \eIf(\textit{\footnotesize \# Change  is too small
              to
              continue.})
            {$\| \delta \| \le \epsilon_2 \|
              (\vec{\phi},\vec{\theta},\mu) \|$} {
				$\text{stop} \leftarrow \text{true}$\;
			}{
				$(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})
                \leftarrow (\vec{\phi},\vec{\theta},\mu) + \delta$
                \textit{\footnotesize \# A trial step}\;
				$\rho \leftarrow (\|
                Z(\vec{\phi},\vec{\theta},\mu)\|^2 - \|
                Z(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})\|^2
                )/(\delta^T(u \delta + g))$ \;
                \eIf(\textit{\footnotesize \# Trial step was good}){$\rho >
                    0$}{
					$(\vec{\phi},\vec{\theta},\mu) \leftarrow
                    (\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})$
                    \textit{\footnotesize \# Update variables}\;
                    Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
                    Eq. (9); $A \leftarrow J^T J$; $g \leftarrow J^T
                    Z(\vec{\phi},\vec{\theta},\mu)$\;
					$ \text{stop} \leftarrow (\|g\|_{\infty} \le
                    \epsilon_1)$ or $(\|
                    Z{(\vec{\phi},\vec{\theta},\mu)}^2 \| \le
                    \epsilon_3)$ \;
                    $v \leftarrow 2$;
					$u \rightarrow u * \max(1/3, 1 - {(2\rho - 1)}^3 )$\;
                } (\textit{\footnotesize \# Trial step was bad}){
                    $v \leftarrow 2 v$;
					$u \leftarrow u v$\;
				}
			}
		 }
	}
	$(\vec{\phi}^*,\vec{\theta}^*,\mu^*) \leftarrow (\vec{\phi},\vec{\theta},\mu)$\;

\label{alg:LM}
\end{algorithm}

Suggested values for the tuning parameters are $\epsilon_1 = \epsilon_2 =
\epsilon_3 = 10^{-15}, \tau = 10^{-3}$ and $k_{\max} = 100$.

The Jacobian matrix $J = {(J_{1}, J_{2}, \ldots, J_N)}^T$ requires the partial derivatives, which are
\begin{equation}
J_t = {(J_{t, \phi_1}, \ldots, J_{t,\phi_p}, J_{t,\theta_1}, \ldots,
  J_{t,\theta_q}, J_{t,\mu})}^T
\end{equation}
Here the last term is present only when \texttt{include\_mean} is
\texttt{True}.
The iteration relations for $J$ are
\begin{eqnarray}
J_{t, \phi_i} &=& \frac{\partial F_t(\phi,\theta)}{\partial \phi_i} =
-\frac{\partial Z_t}{\partial \phi_i} = X_{t-i}-\mu + \sum_{j=1}^q
\theta_j \frac{\partial Z_{t - j}}{\partial \phi_i} = X_{t-i}-\mu - \sum_{j=1}^q
\theta_j J_{t-j,\phi_i}, \label{eq:J1}\\
J_{t, \theta_i}&=&\frac{\partial F_t(\phi,\theta)}{\partial \theta_i} =
-\frac{\partial Z_t}{\partial \theta_i} = Z_{t-i} + \sum_{j =1}^q
\theta_j \frac{\partial Z_{t - j}}{\partial \theta_i} = Z_{t-i} -
\sum_{j=1}^q \theta_j J_{t-j,\theta_i}, \label{eq:J2}\\
J_{t, \mu} &=&\frac{\partial F_t(\phi,\theta)}{\partial \mu} =
-\frac{\partial Z_t}{\partial \mu} = 1 -
\sum_{j=1}^p \phi_j - \sum_{j=1}^q \theta_j \frac{\partial
  Z_{t-j}}{\partial \mu} = 1 - \sum_{j=1}^p \phi_j - \sum_{j=1}^q
\theta_j J_{t-j,\mu}. \label{eq:J3}
\end{eqnarray}
Note that the mean value $\mu$ is considered separately in the above
formulations. When \texttt{include\_mean} is set to \texttt{False}, $\mu$ will be simply
set to 0. Otherwise, $\mu$ will also be estimated together with
$\vec{\phi}$ and $\vec{\theta}$. The initial conditions for the above
equations are
\begin{equation}
J_{t,\phi_i} = J_{t,\theta_j} = J_{t,\mu} = 0 \quad \mbox{for }
t \leq p,\ \mbox{and }i=1,\dots,p; j = 1, \dots, q\ ,
\end{equation}
because we have fixed $Z_t$ for $t\leq p$ to be a constant $0$ in the initial
condition. Note that $J$ is zero not only for $t\leq
0$ but also for $t\leq p$.

@@

Eqs. (\ref{eq:error-terms},~\ref{eq:J1},~\ref{eq:J2},~\ref{eq:J3}) are
difficult to parallelize.
